---
title: "EDA"
author: "Trevor Lynch"
format: html
editor: visual
---

```{r}
library(psych)
library(tidyverse)
```


## Introduction

briefly describes the data and the variables you have to work with (just discuss the ones you’ll investigate/use in your analysis).


describes the purpose of your EDA and ultimate goal of modeling.


## Data
```{r}
diabetes <- read.csv("diabetes_binary_health_indicators_BRFSS2015.csv", header = T)
str(diabetes)

diabetes_tib <- as_tibble(diabetes) |>
  mutate(Diabetes_binary = as.factor(Diabetes_binary),
         HighBP = as.factor(HighBP),
         HighChol = as.factor(HighChol),
         CholCheck = as.factor(CholCheck),
         Smoker = as.factor(Smoker),
         Stroke = as.factor(Stroke),
         HeartDiseaseorAttack = as.factor(HeartDiseaseorAttack),
         PhysActivity = as.factor(PhysActivity),
         Fruits = as.factor(Fruits),
         Veggies = as.factor(Veggies),
         HvyAlcoholConsump = as.factor(HvyAlcoholConsump),
         AnyHealthcare = as.factor(AnyHealthcare),
         NoDocbcCost = as.factor(NoDocbcCost),
         GenHlth = as.factor(GenHlth),
         DiffWalk = as.factor(DiffWalk),
         Sex = as.factor(Sex),
         Age = as.factor(Age),
         Education = as.factor(Education),
         Income = as.factor(Income))

# Check for missing values in the dataset
apply(diabetes_tib, MARGIN = 2, function(x) sum(is.na(x)))
# There are no missing values marked in the dataset

# Produce Frequency tables for all of the factor variables
apply(dplyr::select(diabetes_tib, -c(BMI, MentHlth, PhysHlth)), MARGIN = 2, table)

# Check correlation of continuous variables
diabetes_sub <- dplyr::select(diabetes_tib, c(Diabetes_binary, BMI, MentHlth, PhysHlth))
cor(diabetes_sub, use = "pairwise.complete.obs")

```
Use a relative path to import the data. You likely want to convert a lot of the variables to factors with
meaningful level names.
Check on missingness, etc.
```{r}

```


## Summarizations
```{r}

```


At the end of the document: 
[Click here for the Modeling Page](Modeling.html)